Assessing linkages between watershed nutrient loading and water quality in a subtropical estuary with semiparametric models.

### 2023-05-20 Schramm-2023-04-AS/Lavaca-Manuscript-AS.qmd
init springer nature tex submission for AS

---
title: "Assessing linkages between watershed nutrient loading and water quality in a subtropical estuary with semiparametric models."
author: 
  - name: Michael Schramm
    email: michael.schramm@ag.tamu.edu
    corresponding: true
    affiliations:
      - ref: agrilife
affiliations:
  - id: agrilife
    name: Texas A&M AgriLife Research, Texas Water Resources Institute
    city: College Station
    state: Texas
    country: United States of America
format: 
  docx:
    number-sections: false
    reference-doc: format.docx
crossref:
  fig-prefix: Fig   # (default is "Figure")
  tbl-prefix: Table    # (default is "Table")
  title-delim: .     # (default is ":")
  ref-hyperlink: false # (default is true)
bibliography: mybibfile.bib
csl: https://www.zotero.org/styles/plos?source=1
filters: 
  - authors-affiliations.lua
keep-md: true
---

<!--
TODO:
- find right csl format
- fix cross refs
- fix equations
- bold captions(?)
- fix symbols, subscripts, superscripts in tables
 List corresponding author’s initials in 
parentheses after the email address.
make intial field? for corresponding author?
- use imagemagick to convert png to tiff?
-->

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(targets)
library(knitr)
library(tidyverse)
library(kableExtra)
library(patchwork)
library(ggrepel)
library(rcartocolor)
library(twriTemplates)
library(ggspatial)
library(ragg)
library(flextable)
library(modelsummary)
library(scico)
library(ggridges)
library(ggtext)
library(lubridate)
library(units)
library(gratia)
library(gt)

@@

## this sets our default code chunk options
knitr::opts_chunk$set(echo = FALSE,
                      ## figure chunk options
                      dev = "CairoPS",
                      dpi = 600,
                      fig.align = "center",
                      out.width = "100%")

@@

# Abstract
Lavaca Bay is a small secondary embayment on the Texas coast that is
displaying early signals of water quality degradation. This
study applied a semiparametric approach to assess both watershed nutrient loads 
and the responses in estuary water quality to nutrient loading and streamflow. 
Cross-validation indicated that, despite data constraints, semiparametric models 
performed well at nutrient load prediction. Based on these models, delivered
annual nutrient loads varied substantially from year to year. In contrast,
minimal changes in calculated flow-normalized loads indicate that changes in 
nutrient loading were driven by natural variation in precipitation and runoff as
opposed to changes in discharge or nonpoint source management efforts.
Models did not identify significant long-term changes within Lavaca Bay for dissolved 
oxygen or chlorophyll-*a*. However, site specific long-term increases in both
organic and inorganic nitrogen are concerning. Further analysis found freshwater
inflow was a strong driver of nutrient and chlorophyll-*a* concentrations within
Lavaca Bay but there was no evidence that changes in nutrient loading explained
variation in dissolved oxygen or chlorophyll-*a* concentrations. In addition to
providing a baseline assessment of watershed nutrient loading and water quality
responses in the Lavaca Bay watershed, this study provides methodological
support for the use of semiparametric methods in load regression models and 
estuary assessments.

@@

Like many coastal areas globally, the coastal watersheds along the Texas Gulf coast 
are facing pressures from increasing population, increases in point source 
and non-point source pollution and alterations to freshwater flows that degrade
water quality in downstream estuaries [@bricker_effects_2008; @kennicuttWaterQualityGulf2017; @bugica_water_2020]. Despite these increasing pressures, national scale 
assessments have classified coastal estuaries in Texas as moderate or lower for
exhibiting eutrophic conditions [@bricker_effects_2008]. However, a suite of 
recent studies indicate that estuary water quality dynamics in both 
agriculturally dominated and urban watersheds within Texas are in fact 
expressing conditions that are increasingly conducive to algal blooms and 
eutrophication [@wetzWaterQualityDynamics2016; @wetz_exceptionally_2017; 
@bugica_water_2020; @chinPhytoplanktonBiomassCommunity2022]. With identification
of localized areas of estuary water quality concern along the Texas coast 
[@bugica_water_2020], localized studies are being prioritized to better 
inform management actions.

This project provides an assessment of nutrient loading and water quality
responses in Lavaca Bay, Texas. Lavaca Bay is a secondary bay in the larger
Matagorda Bay system located roughly halfway between Houston, Texas and Corpus
Christi, Texas. Lavaca Bay faces substantial challenges associated with
legacy contamination but general water quality parameters such as dissolved oxygen 
(DO), nutrients, and biological parameters have been well within state
water quality standards. However, long-term declines of benthic 
fauna abundance, biomass, and diversity in Lavaca Bay primarily
linked to reductions in freshwater inflows and changes in estuary salinity 
[@beserespollackLongtermTrendsResponse2011; @palmerImpactsDroughtsLow2015;
@montagnaAssessmentRelationshipFreshwater2020] are a concern to local 
stakeholders. Recent water quality assessments identified monotonic increases in total 
phosphorus (TP), orthophosphate, total Kjeldahl nitrogen (TKN), and
chlorophyll-*a* at sites within Lavaca Bay [@bugica_water_2020]. Although 
long-term changes in DO concentrations were not identified, the trends in 
nutrient concentrations are concerning due to the role of nitrogen as a 
limiting factor for primary production in many Texas estuaries 
[@gardnerNitrogenFixationDissimilatory2006; @houTransformationFateNitrate2012; @doradoUnderstandingInteractionsFreshwater2015; @paudelRelationshipSuspendedSolids2019; @wetz_exceptionally_2017] and the 
ramifications that changes in nitrogen loadings could have for 
productivity and eutrophication in Lavaca Bay.

There are ongoing efforts between local, state, and federal agencies to 
address water quality impairments in the freshwater portions of the 
Lavaca Bay watershed [@jainTechnicalSupportDocument2021; @schramm_lavaca_2018; @bertholdDirectMailingEducation2021]. However, at a statewide scale, these
approaches have shown limited success and emphasize a need for improved efforts
at assessing and linking management actions with downstream water quality to
better identify  and replicate effective management actions across the state
[@schrammTotalMaximumDaily2022]. The identification and communication of 
changes and trends in water quality is complicated by the fact that trends are 
often non-linear and confounded by precipitation and runoff that 
hinder traditional analysis [@wazniakLinkingWaterQuality2007; 
@lloydMethodsDetectingChange2014]. To provide
actionable information for resource managers, water quality conditions must be 
evaluated relative to changes in natural environmental drivers to 
better understand  and manage potential anthropogenic effects.
This study utilizes semiparemetric methods to develop estimates of delivered 
and flow-normalized nutrient loads and assess changes in loads delivered to 
Lavaca Bay. The study also assesses the response of water quality parameters in
Lavaca Bay over time and in response to freshwater inflow controlled
for seasonality and to watershed nutrient loads that are controlled for
environmentally driven variation.

@@

Water quality sample data for the two freshwater and three estuary locations were 
obtained from the Texas Commission on Environmental Quality (TCEQ) Surface Water 
Quality Monitoring Information System. Data submitted through the system 
are required to be collected under Quality Assurance Project Plans and lab 
method procedures outlined by the TCEQ's procedures manual. 
The QAPP and procedures manuals ensure the consistent collection and laboratory 
methods are applied between samples collected by different entities and under 
different projects. 
All sites had varying lengths of and availability of data.
For freshwater locations, TP from January 2000 through December 2020 and 
nitrate-nitrogen (NO~3~) data from January 2005 through December 2020 were 
downloaded (@tbl-table1). Less than 5-years of total
nitrogen and TKN concentration data were available at the freshwater sites and
deemed insufficient to develop load estimation models 
[@horowitzEvaluationSedimentRating2003; @snelderEstimationCatchmentNutrient2017]. 
The three estuary sites included an upper Lavaca Bay site near the outlet of the
Lavaca River system (TCEQ-13563), a mid-Lavaca Bay site (TCEQ-13383), and the 
lower Lavaca Bay site near the mouth of the Bay (TCEQ-13384). 
For estuary locations, we obtained data for TP, Nitrite+Nitrate (NO*~x~*), TKN, 
chlorophyll-*a*, and DO concentrations from January 2005 through December 2020 
(@tbl-table2).

```{r}
#| label: tbl-table1
#| tbl-cap: "Summary of gauged streamflow and freshwater water quality samples between January 1, 2000 and December 31, 2020."
#| warning: false
df <- tar_read(model_data, store = store)

@@

datasummary(`Station ID` * (`TP (mg/L)` + `NO3 (mg/L)` + `Mean Daily Streamflow (cfs)`) ~
               Mean*Arguments(na.rm=TRUE) + SD*Arguments(na.rm=TRUE) + N,
            data = df, output = "kableExtra", format = "pipe") 
```

```{r}
#| label: tbl-table2
#| warning: false
#| tbl-cap: "Summary of estuary water quality samples collected between January 1, 2005 and December 31, 2020."
df <- tar_read(estuary_model_data, store = store)

@@

 modelsummary::datasummary(`Station ID` * (`TP (mg/L)`+
                            `NOx (mg/L)` +
                            `TKN (mg/L)` +
                            `Chlorophyll-a (ug/L)` +
                            `DO (mg/L)`) ~
                            Mean*Arguments(na.rm=TRUE) + SD*Arguments(na.rm=TRUE) + N,
                          data = df, output = "kableExtra", format = "pipe") #|>
  #  flextable::autofit()
   # mutate(`Station ID` = case_when(
   #   `Station ID` == "" ~ NA_character_,
   #   `Station ID` != "" ~ `Station ID`
   #   )) |>
   # tidyr::fill(`Station ID`, .direction = "down") |> 
   # gt::gt() |> 
   # gt::as_word()

@@

Estimates of NO~3~ and TP loads at the Lavaca River (USGS-08164000) and the 
outlet of Lake Texana on the Navidad River (USGS-08164525) were developed using 
Generalized Additive Models (GAMs) relating nutrient concentration to river 
discharge, season, and time.
Separate models were fit at each station for each parameter and used to predict
nutrient concentrations for each day in the study period. GAMs can be specified
in a functionally similar manner to the commonly used LOADEST 
[@cohn_validity_1992] or WRTDS 
[@hirsch_weighted_2010] regression models and have been shown to
produce reliable estimates of nutrient and sediment loadings 
[@wangLoadEstimationUncertainties2011; @kroonRiverLoadsSuspended2012; 
@kuhnert_quantifying_2012; @robson_prediction_2015-1; 
@hagemannEstimatingNutrientOrganic2016; @mcdowell_implications_2021; 
@biagi_novel_2022]. GAMs are a semiparametric extension of generalized linear 
models where the linear predictor is represented as the sum of multiple unknown 
smooth functions and parametric linear predictors [@wood_fast_2011]. Although 
the underlying parameter estimation procedure of GAMs is substantially different 
than WRTDS, both the functional form and results are demonstrated to be similar [@beckNumericalQualitativeContrasts2017]. GAMs were chosen over other regression-based
approached for use in this study due to; (1) the ability to easily explore and 
incorporate different model terms; (2) the incorporation of non-liner smooth
functions that do not require explicit a priori knowledge of the expected
shape; and (3) inclusion of a link function that related the expected value of
the response to linear predictors thus avoiding unneeded data transformations
and bias corrections.

GAMs were fit using the *mgcv* package in R which makes available multiple 
types of smooth functions with automatic smoothness selection [@wood_fast_2011]. 
The general form of the model related NO~3~ or TP concentration to a long term
tend, season, streamflow, and two different antecedent discharge terms:

$$
g(\mu) = \alpha + f_1(ddate) + f_2(yday) + f_3(log1p(Q)) + f_4(ma) + f_5(fa)  
y \sim \mathcal{N}(\mu,\,\sigma^{2})
$${#eq-eq1}

@@

Moving average (*ma*) is an exponentially smoothed moving average that attempts 
to incorporate the influence of prior streamflow events on concentration at the 
current time period [@wangLoadEstimationUncertainties2011;  
@kuhnert_quantifying_2012; @zhang_improving_2017]:

$$
ma(\delta) = d{\kappa_{i-1}}+(1-\delta)\hat{q}_{i-1}\quad\text{and}\quad \kappa_{i}=\sum_{m=1}^{i}\hat{Q}_m
$${#eq-eq2}

@@

Flow anomaly (*fa*) is a unitless term that represents how wet or dry the 
current time period is from a previous time period [@vecchia_trends_2009; 
@zhang_improving_2017]. Long-term flow anomaly (*ltfa*) is the streamflow over 
the previous year relative to the entire period and calculated as described as
[@zhang_improving_2017]:

@@

and the short-term flow anomaly (*stfa*) calculated as the current day flow 
compared to the preceding 1-month streamflow:

@@

The calculation of model terms for the Lake Texana site were slightly modified
because daily loads are not a function of natural stream flow processes alone,
but of dam releases and nutrient concentrations at the discharge point of the 
lake. *Q*, *ma*, and *fa* terms were calculated based on total gaged inflow from
the 4 major tributaries to the lake. Thin-plate regression splines were used 
for *ddate*, *log1p(Q)*, *fa*, and *ma*. A cyclic cubic regression spline was 
used for *yday* to ensure the ends of the spline match 
(day 1 and day 366 are expected to match). First order penalties were applied 
to the smooths of flow-based variables which penalize departures from a 
flat function to help constrain extrapolations for high flow measurements.

@@

Daily loads were estimated as the predicted concentration multiplied by the 
daily streamflow. For the Navidad River (USGS-08164525) site, daily loads at the 
dam were calculated 
from the discrete daily concentration at the discharge point of the lake and 
corresponding reported daily discharge from the dam.
Flow-normalized loads were estimated similar to WRTDS by setting flow-based 
covariates on each day of the year equal to each of the historical values for
that day of the year over the study period [@hirsch_weighted_2010]. The 
flow-normalized estimate was calculated as the mean of all the predictions for 
each day considering all possible flow values.
Standard deviations and credible intervals were obtained by 
drawing samples from the multivariate normal posterior distribution of the 
fitted GAM [@woodConfidenceIntervalsGeneralized2006; 
@marraCoveragePropertiesConfidence2012; @mcdowell_implications_2021].
Uncertainty in loads were calculated as 90% credible intervals estimated by 
drawing 1000 realizations of parameter estimates from the multivariate normal 
posterior distribution of the model parameters. GAM performance was evaluated 
using repeated 5-fold cross validation  [@burmanComparativeStudyOrdinary1989] 
and average Nash-Sutcliffe Efficiency (NSE), r^2^ and percent bias (PBIAS) 
metrics across folds were calculated for each model.

@@

To test if changes in freshwater inflow and nutrient loading had explanatory
effect on changes in estuary water quality a series of GAM models were fit at
each site relating parameter concentration to temporal trends, inflow, and 
nutrient loads [@murphyNutrientImprovementsChesapeake2022]:

$$
g(\mu) = \alpha + f_1(ddate) + f_2(yday) y \sim \Gamma(\mu,\lambda)
$${#eq-eq5}

$$
g(\mu) = \alpha + f_1(ddate) + f_2(yday) + f_3(Q) y \sim \Gamma(\mu,\lambda)
$${#eq-eq6}

$$
g(\mu) = \alpha + f_1(ddate) + f_2(yday) + f_3(Q) + f_4(Load) y \sim \Gamma(\mu,\lambda)
$${#eq-eq7}

@@

```{r}
#| label: tbl-table3
#| tbl-cap: "Set of GAM models specified for each water quality parameter response."
#| warning: false
df <- tibble(
  `Water Quality Response Parameter` = c(rep("TP", 3), rep("NOx", 3),
               rep("Chlorophyll-a", 3), rep("Dissolved Oxygen", 3),
               rep("TKN", 2)),
  Model = c("Temporal", "Flow", "Flow+Load",
            "Temporal", "Flow", "Flow+Load",
            "Temporal", "Flow", "Flow+Load",
            "Temporal", "Flow", "Flow+Load",
            "Temporal", "Flow"),
  `Model Terms` = c("s(ddate) + s(yday)",
                    "s(ddate) + s(yday) + s(Q)",
                    "s(ddate) + s(yday) + s(Q) + s(TP Load)",
                    "s(ddate) + s(yday)",
                    "s(ddate) + s(yday) + s(Q)",
                    "s(ddate) + s(yday) + s(Q) + s(NO3 Load)",
                    "s(ddate) + s(yday)",
                    "s(ddate) + s(yday) + s(Q)",
                    "s(ddate) + s(yday) + s(Q) + s(TP Load) + s(NO3 Load)",
                    "s(ddate) + s(yday)",
                    "s(ddate) + s(yday) + s(Q)",
                    "s(ddate) + s(yday) + s(Q) + s(TP Load) + s(NO3  Load)",
                    "s(ddate) + s(yday)",
                    "s(ddate) + s(yday) + s(Q)")
)

kbl(df, format = "pipe")

>>>>

---
title: "Assessing linkages between watershed nutrient loading and water quality in a subtropical estuary with semiparametric models."
format:
  wordcount-docx:
    toc: false
    number-sections: false
    reference-doc: format.docx
author: 
  - name: Michael Schramm
    email: michael.schramm@ag.tamu.edu
    corresponding: true
    affiliations:
      - ref: agrilife
affiliations:
  - id: agrilife
    name: Texas A&M AgriLife Research, Texas Water Resources Institute
    city: College Station
    state: Texas
    country: United States of America
crossref:
  fig-prefix: Fig   # (default is "Figure")
  tbl-prefix: Table    # (default is "Table")
  title-delim: .     # (default is ":")
  ref-hyperlink: false # (default is true)
  eq-prefix: Eq
bibliography: mybib.yaml
csl: https://www.zotero.org/styles/springer-basic-author-date-no-et-al?source=1
filters: 
  - authors-affiliations.lua
keep-md: true
cite-method: citeproc
link-bibliography: false
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(targets)
library(knitr)
library(tidyverse)
library(kableExtra)
library(patchwork)
library(ggrepel)
library(rcartocolor)
library(twriTemplates)
library(ggspatial)
library(ragg)
library(flextable)
library(modelsummary)
library(scico)
library(ggridges)
library(ggtext)
library(lubridate)
library(units)
library(gratia)
library(gt)
library(officer)

@@

## this sets our default code chunk options
knitr::opts_chunk$set(echo = FALSE,
                      ## figure chunk options
                      dev = "ragg_tiff",
                      dpi = 300,
                      fig.align = "center",
                      fig.ext = "tiff",
                      out.width = "100%")

@@

# Abstract
Lavaca Bay is a small secondary embayment on the Texas coast that is
displaying early signals of water quality degradation. This
study applied a semiparametric approach to assess both watershed nutrient loads 
and the responses in estuary water quality to nutrient loading and streamflow. 
Cross-validation indicated that, despite data constraints, semiparametric models 
performed well at nutrient load prediction. Based on these models, delivered
annual nutrient loads varied substantially from year to year. In contrast,
minimal changes in calculated flow-normalized loads indicate that changes in 
nutrient loading were driven by natural variation in precipitation and runoff as
opposed to changes in nonpoint sources.
Estuary water quality models did not identify significant long-term changes 
within Lavaca Bay for dissolved oxygen or chlorophyll-*a*. However, site specific 
long-term increases in both organic and inorganic nitrogen are concerning. 
Further analysis found freshwater inflow was a strong driver of nutrient and 
chlorophyll-*a* concentrations within Lavaca Bay but there was no evidence that 
changes in nutrient loading explained variation in dissolved oxygen or 
chlorophyll-*a* concentrations. In addition to providing a baseline assessment 
of watershed nutrient loading and water quality responses in the Lavaca Bay 
watershed, this study provides methodological support for the use of 
semiparametric methods in load regression models and estuary assessments.

@@

Like many coastal areas globally, the coastal watersheds along the Texas Gulf coast 
are facing pressures from increasing population, increases in point source 
and non-point source pollution and alterations to freshwater flows that degrade
water quality in downstream estuaries [@bricker_effects_2008; @kennicuttWaterQualityGulf2017; @bugica_water_2020]. 
Despite these escalating pressures, national scale assessments have classified 
coastal estuaries in Texas as moderate or low risk for eutrophic conditions 
[@bricker_effects_2008]. However, a suite of recent studies indicates that 
estuary water quality dynamics in both agriculturally dominated and urban 
watersheds within Texas are in fact expressing conditions that are increasingly 
conducive to algal blooms and eutrophication [@wetzWaterQualityDynamics2016; 
@wetz_exceptionally_2017; @bugica_water_2020; @chinPhytoplanktonBiomassCommunity2022]. 
With identification of localized areas of estuary water quality concern along 
the Texas coast [@bugica_water_2020], localized studies are being prioritized 
to better inform management actions.

This project aims to provide an assessment of nutrient loading and water quality
responses in Lavaca Bay, Texas. Lavaca Bay is a secondary bay in the larger
Matagorda Bay system located roughly halfway between Houston, Texas and Corpus
Christi, Texas. Lavaca Bay faces substantial challenges associated with
legacy contamination but general water quality parameters such as dissolved oxygen 
(DO), nutrients, and biological parameters have been well within state
water quality standards. More recently identified long-term declines 
in abundance, biomass, and diversity of benthic fauna in Lavaca Bay have been 
linked to reductions in freshwater inflows and changes in estuary salinity 
[@beserespollackLongtermTrendsResponse2011; @palmerImpactsDroughtsLow2015;
@montagnaAssessmentRelationshipFreshwater2020] and are a concern to local 
stakeholders. Water quality assessments identified monotonic increases 
in total phosphorus (TP), orthophosphate, total Kjeldahl nitrogen (TKN), and
chlorophyll-*a* at sites within Lavaca Bay [@bugica_water_2020]. Although 
long-term changes in DO concentrations were not identified, the trends in 
nutrient concentrations are concerning due to the role of nitrogen as a 
limiting factor for primary production in many Texas estuaries 
[@gardnerNitrogenFixationDissimilatory2006; @houTransformationFateNitrate2012;
@doradoUnderstandingInteractionsFreshwater2015; 
@paudelRelationshipSuspendedSolids2019; @wetz_exceptionally_2017] and the 
ramifications that changes in nitrogen loadings could have for 
productivity and eutrophication in Lavaca Bay.

There are ongoing efforts between local, state, and federal agencies to 
address water quality impairments in the freshwater portions of the 
Lavaca Bay watershed [@jainTechnicalSupportDocument2021; 
@schrammLavacaRiverWatershed2018; @bertholdDirectMailingEducation2021]. 
However, at a statewide scale, these approaches have shown limited success and 
emphasize a need for improved efforts at assessing and linking management 
actions with downstream water quality to identify and replicate effective 
management actions across the state [@schrammTotalMaximumDaily2022]. The 
identification and communication of changes and trends in water quality is 
complicated by the fact that trends are often non-linear and confounded by 
precipitation and runoff that hinder traditional analysis 
[@wazniakLinkingWaterQuality2007; @lloydMethodsDetectingChange2014]. 
The development and application of statistical methods such as Weighted
Regressions on Time, Discharge and Season [WRTDS, @hirschWeightedRegressionsTime2010] 
and Generalized Additive Models [GAMs, @woodFastStableRestricted2011] has 
provided effective tools for researchers to quantify and
communicate non-linear changes in river and estuary pollutant loadings.

WRTDS calculates a time series of in stream concentrations or loads 
(daily, monthly, or annually) and flow-normalized estimates of concentrations and loads
using locally weighted regresssion for unique combinations of time, discharge, and season. 
WRTDS has been widely used to assess and identify trends in riverine nutrients 
[@oelsner_recent_2019; @stackpooleLongTermMississippi2021],
chlorides [@stetsIncreasingChlorideRivers2018], and other pollutants of concern 
[@shodaWaterqualityTrendsRivers2019]. WRTDS has also been succesfully adapted to 
assess trends in estuarine water quality concentrations [@beckFourDecadesWater2018].

While WRTDS is a statistical approach developed specifically for water quality applications, 
GAMs are a broadly applicable statistical method. GAMs are a semiparametric 
extension of generalized linear models where the linear predictor is represented
as the sum of multiple unknown smooth functions and parametric linear predictors
[@woodFastStableRestricted2011]. Although the underlying parameter estimation 
procedure of GAMs is substantially different than WRTDS, both the functional 
form and results have been demonstrated to be similar 
[@beckNumericalQualitativeContrasts2017]. Water quality applications of GAMs
include river and catchment nutrient concentration and load models 
[@wangLoadEstimationUncertainties2011;@kroonRiverLoadsSuspended2012;
@kuhnertQuantifyingTotalSuspended2012;@robson_prediction_2015-1;
@hagemannEstimatingNutrientOrganic2016;@mcdowell_implications_2021;
@biagi_novel_2022]. GAMs can also be used to identify non-linear temporal trends
(including flow-normalized trends) in pollutant concentrations and loads 
[@beckNumericalQualitativeContrasts2017; @murphyGeneralizedAdditiveModel2019]. 
Recently GAMs have also been used to link water quality responses in 
receiving water bodies to changes in nutrient inputs 
[@murphyNutrientImprovementsChesapeake2022]. @beckNumericalQualitativeContrasts2017
provides a substantial discussion on the differences (and similarities) between 
GAMs and WRTDS for water quality applications.

To provide actionable information for resource managers in Lavaca Bay, 
water quality conditions must be evaluated relative to changes in natural 
environmental drivers to better understand and manage potential human impacts.
This study utilizes GAMs to develop estimates of delivered 
and flow-normalized nutrient loads and assess changes in loads delivered to 
Lavaca Bay. GAMs were chosen over other regression-based
approached for use in this study due to; (1) the ability to easily explore and 
incorporate different model terms; (2) the incorporation of non-liner smooth
functions that do not require explicit a priori knowledge of the expected
shape; and (3) inclusion of a link function that related the expected value of
the response to linear predictors thus avoiding unneeded data transformations
and bias corrections. The study also assesses the response of water quality parameters in
Lavaca Bay over time and in response to freshwater inflow controlled
for seasonality and to watershed nutrient loads that are controlled for
environmentally driven variation.

@@

Water quality sample data for the two freshwater and three estuary locations were 
obtained from the Texas Commission on Environmental Quality (TCEQ) Surface Water 
Quality Monitoring Information System. Data submitted through the system 
are required to be collected under Quality Assurance Project Plans and lab 
method procedures outlined by the TCEQ's procedures manual. 
The QAPP and procedures manuals ensure the consistent collection and laboratory 
methods are applied between samples collected by different entities and under 
different projects. 
All sites had varying lengths of and availability of data.
For freshwater locations, TP from January 2000 through December 2020 and 
nitrate-nitrogen (NO~3~) data from January 2005 through December 2020 were 
downloaded (`#r run_reference("table1")`). Less than 5-years of total
nitrogen and TKN concentration data were available at the freshwater sites and
deemed insufficient to develop load estimation models 
[@horowitzEvaluationSedimentRating2003; @snelderEstimationCatchmentNutrient2017]. 
The three estuary sites included an upper Lavaca Bay site near the outlet of the
Lavaca River system (TCEQ-13563), a mid-Lavaca Bay site (TCEQ-13383), and the 
lower Lavaca Bay site near the mouth of the Bay (TCEQ-13384). 
For estuary locations, we obtained data for TP, Nitrite+Nitrate (NO*~x~*), TKN, 
chlorophyll-*a*, and DO concentrations from January 2005 through December 2020 
(@tbl-table2).

```{r}

df <- tar_read(model_data, store = store)

@@

datasummary(`Station ID` * (`TP (mg/L)` + `NO3 (mg/L)` + `Mean Daily Streamflow (cfs)`) ~
               Mean*Arguments(na.rm=TRUE) + SD*Arguments(na.rm=TRUE) + N,
            data = df, output = "kableExtra", format = "simple")

# datasummary(`Station ID` * (`TP (mg/L)` + `NO3 (mg/L)` + `Mean Daily Streamflow (cfs)`) ~
#                Mean*Arguments(na.rm=TRUE) + SD*Arguments(na.rm=TRUE) + N,
#             data = df, output = "flextable") |> 
#   set_caption(caption = "Summary of gauged streamflow and freshwater water quality samples between January 1, 2000 and December 31, 2020.",
#               autonum = run_autonum(seq_id = "tab", bkm = "table1")) |> 
#   style(pr_t = fp_text_default(font.size = 9), part = "body") |>
#   style(pr_t = fp_text_default(font.size = 9, bold = TRUE), part = "header") |>
#   set_table_properties(layout = "autofit") |>
#   padding(padding.top = 0, padding.bottom = 0, part = "body")
```

```{r}
#| warning: false

df <- tar_read(estuary_model_data, store = store)

@@

modelsummary::datasummary(`Station ID` * (`TP (mg/L)`+
                                            `NOx (mg/L)` +
                                            `TKN (mg/L)` +
                                            `Chlorophyll-a (ug/L)` +
                                            `DO (mg/L)`) ~
                            Mean*Arguments(na.rm=TRUE) + SD*Arguments(na.rm=TRUE) + N,
                          data = df, output = "flextable") |> 
  set_caption(caption = "Summary of estuary water quality samples collected between January 1, 2005 and December 31, 2020.",
              autonum = run_autonum(seq_id = "tab", bkm = "table2")) |> 
  style(pr_t = fp_text_default(font.size = 9), part = "body") |>
  style(pr_t = fp_text_default(font.size = 9, bold = TRUE), part = "header") |>
  set_table_properties(layout = "autofit") |>
  padding(padding.top = 0, padding.bottom = 0, part = "body")

 # modelsummary::datasummary(`Station ID` * (`TP (mg/L)`+
 #                            `NOx (mg/L)` +
 #                            `TKN (mg/L)` +
 #                            `Chlorophyll-a (ug/L)` +
 #                            `DO (mg/L)`) ~
 #                            Mean*Arguments(na.rm=TRUE) + SD*Arguments(na.rm=TRUE) + N,
 #                          data = df, output = "kableExtra", format = "pipe") #|>
  #  flextable::autofit()
   # mutate(`Station ID` = case_when(
   #   `Station ID` == "" ~ NA_character_,
   #   `Station ID` != "" ~ `Station ID`
   #   )) |>
   # tidyr::fill(`Station ID`, .direction = "down") |> 
   # gt::gt() |> 
   # gt::as_word()

@@

Estimates of NO~3~ and TP loads at the Lavaca River (USGS-08164000) and the 
outlet of Lake Texana on the Navidad River (USGS-08164525) were developed using 
GAMs relating nutrient concentration to river discharge, season, and time.
Separate models were fit at each station for each parameter and used to predict
nutrient concentrations for each day in the study period. GAMs were fit using 
the *mgcv* package in R which makes available multiple 
types of smooth functions with automatic smoothness selection [@woodFastStableRestricted2011]. 
The general form of the model related NO~3~ or TP concentration to a long term
tend, season, streamflow, and two different antecedent discharge terms, 
shown in [@eq-eq1]:

$$
\begin{align*}
g(\mu) &= \alpha + f_1(ddate) + f_2(yday) + f_3(log1p(Q)) + f_4(ma) + f_5(fa)  \\
y &\sim \mathcal{N}(\mu,\,\sigma^{2})
\end{align*}
$${#eq-eq1}

@@

Moving average (*ma*) is an exponentially smoothed moving average that attempts 
to incorporate the influence of prior streamflow events on concentration at the 
current time period [@wangLoadEstimationUncertainties2011;  
@kuhnertQuantifyingTotalSuspended2012; @zhang_improving_2017], using [@eq-eq2]:

$$
\begin{align*}
ma(\delta) &= d{\kappa_{i-1}}+(1-\delta)\hat{q}_{i-1},\\
\kappa_{i} &=\sum_{m=1}^{i}\hat{Q}_m,
\end{align*}
$${#eq-eq2}

@@

Flow anomaly (*fa*) is a unitless term that represents how wet or dry the 
current time period is from a previous time period [@vecchiaTrendsConcentrationsUse2009; 
@zhang_improving_2017]. Long-term flow anomaly (*ltfa*) is the streamflow over 
the previous year relative to the entire period [@zhang_improving_2017] and 
calculated using [@eq-eq3]:

@@

and the short-term flow anomaly (*stfa*) calculated as the current day flow 
compared to the preceding 1-month streamflow using [@eq-eq4]:

@@

The calculation of model terms for the Lake Texana site were modified
because daily loads are not a function of natural stream flow processes alone,
but of dam releases and nutrient concentrations at the discharge point of the 
lake. *Q*, *ma*, and *fa* terms were calculated based on total gaged inflow from
the 4 major tributaries to the lake. Thin-plate regression splines were used 
for *ddate*, *log1p(Q)*, *fa*, and *ma*. A cyclic cubic regression spline was 
used for *yday* to ensure the ends of the spline match 
(day 1 and day 366 are expected to match). First order penalties were applied 
to the smooths of flow-based variables which penalize departures from a 
flat function to help constrain extrapolations for high flow measurements.

@@

Daily loads were estimated as the predicted concentration multiplied by the 
daily streamflow. For the Navidad River (USGS-08164525) site, daily loads at the 
dam were calculated 
from the discrete daily concentration at the discharge point of the lake and 
corresponding reported daily discharge from the dam.
Flow-normalized loads were estimated similar to WRTDS by setting flow-based 
covariates on each day of the year equal to each of the historical values for
that day of the year over the study period [@hirschWeightedRegressionsTime2010]. The 
flow-normalized estimate was calculated as the mean of all the predictions for 
each day considering all possible flow values.
Standard deviations and credible intervals were obtained by 
drawing samples from the multivariate normal posterior distribution of the 
fitted GAM [@woodConfidenceIntervalsGeneralized2006; 
@marraCoveragePropertiesConfidence2012; @mcdowell_implications_2021].
Uncertainty in loads were calculated as 90% credible intervals estimated by 
drawing 1000 realizations of parameter estimates from the multivariate normal 
posterior distribution of the model parameters. GAM performance was evaluated 
using repeated 5-fold cross validation  [@burmanComparativeStudyOrdinary1989] 
and average Nash-Sutcliffe Efficiency (NSE), Pearson sample correlation (*r*) 
and percent bias (PBIAS) metrics across folds were calculated for each model.

@@

To test if changes in freshwater inflow and nutrient loading had explanatory
effect on changes in estuary water quality a series of GAM models were fit at
each site relating parameter concentration to temporal trends ([@eq-eq5]), 
temporal trends and inflow (@eq-eq6), and temporal trends, inflow, and
nutrient loads (@eq-eq7):

$$
\begin{align*}
g(\mu) &= \alpha + f_1(ddate) + f_2(yday)\\
y &\sim \Gamma(\mu,\lambda)
\end{align*}
$${#eq-eq5}

$$
\begin{align*}
g(\mu) &= \alpha + f_1(ddate) + f_2(yday) + f_3(Q) \\
y &\sim \Gamma(\mu,\lambda)
\end{align*}
$${#eq-eq6}

$$
\begin{align*}
g(\mu) &= \alpha + f_1(ddate) + f_2(yday) + f_3(Q) + f_4(Load) \\
y &\sim \Gamma(\mu,\lambda)
\end{align*}
$${#eq-eq7}

@@

```{r}
#| warning: false
df <- tibble(
  `Water Quality Response Parameter` = c(rep("TP", 3), rep("NOx", 3),
               rep("Chlorophyll-a", 3), rep("Dissolved Oxygen", 3),
               rep("TKN", 2)),
  Model = c("Temporal", "Flow", "Flow+Load",
            "Temporal", "Flow", "Flow+Load",
            "Temporal", "Flow", "Flow+Load",
            "Temporal", "Flow", "Flow+Load",
            "Temporal", "Flow"),
  `Model Terms` = c("s(ddate) + s(yday)",
                    "s(ddate) + s(yday) + s(Q)",
                    "s(ddate) + s(yday) + s(Q) + s(TP Load)",
                    "s(ddate) + s(yday)",
                    "s(ddate) + s(yday) + s(Q)",
                    "s(ddate) + s(yday) + s(Q) + s(NO3 Load)",
                    "s(ddate) + s(yday)",
                    "s(ddate) + s(yday) + s(Q)",
                    "s(ddate) + s(yday) + s(Q) + s(TP Load) + s(NO3 Load)",
                    "s(ddate) + s(yday)",
                    "s(ddate) + s(yday) + s(Q)",
                    "s(ddate) + s(yday) + s(Q) + s(TP Load) + s(NO3  Load)",
                    "s(ddate) + s(yday)",
                    "s(ddate) + s(yday) + s(Q)")
)

# kbl(df, format = "pipe")
